Greenhouse gas emissions model

### 2021-08-23 210617_GHG_model.Rmd
should be ok now for individual animals - next up is modeling an entire population. This will probably just require a loop within the function over all the animal types. In the input table, we'd have to specify the number of animals per type. The feeding system may have to stay fixed (i.e. we'll then be doing a farm-scale analysis, assuming that all animals are in the same feeding system).

The formulas come from Chapter 10 of the *2006 IPCC Guidelines for National Greenhouse Gas Inventories* [@dong_ipcc_2006]. The document can be found on [the IPCC website](https://www.ipcc-nggip.iges.or.jp/public/2006gl/pdf/4_Volume4/V4_10_Ch10_Livestock.pdf). The chapter is part of the book *Good Practice Guidance and Uncertainty Management in National Greenhouse Gas Inventories* [@ipcc_good_2021] , also on the [the IPCC website](https://www.ipcc-nggip.iges.or.jp/public/gp/english/). Many variables are based on the work done in *Variation in the carbon footprint of milk production on smallholder dairy farms in central Kenya* [@wilkes_variation_2020] and in *Methods and guidance to support MRV of livestock emissions* [@wilkes_methods_2019].

@@

# We start with the net energy for maintenance (NE_m; in MJ/day)
# This includes a maintenance coefficient, for which different values are recommended depending on animal type
NE_m<-maintenaince_coefficient_Cfi*(live_weight_LW^0.75)

# The energy use for specific activities is calculated by multiplying NE_m by an activity coefficient (C_activity).

@@

#Net energy for growth (NE_growth; MJ/day):
# here, C is an animal type coefficient
# ### CW Note #### Assuming 'C' is C_activity or should it have been animal_type
C <- C_activity
# C <- animal_type
NE_growth<-4.18*((0.0635*478/(C*FSBW))^0.75*
                   (0.96*SWG)^1.097)

# FSBW: final shrunk body weight in kg - this is the actual final body weight multiplied by 0.96 (to subtract the last meal)

##### EL CW Note ####
# SWG is the "Shrunk Weight Gain" in kg per day - we're not sure what exactly this means.

##### EL CW Note #### 
#This used to be the following equation, which we didn't find a source for:
#NE_growth<-22.02*(((live_weight_LW/(growth_coefficient_C*mature_weight_MW))^0.75)*(weight_gain_WG^1.097))

##### EL CW Note #### 
#The code doesn't include net energy mobilization by animals that lose weight (p. 314)

@@

#Draft animal services are not accounted for

#Energy need during pregnancy 
NE_pregnancy<-0.10*NE_m

##### EL CW Note #### 
#in Joshua's equation the 0.10 was a variable called pregnancy_coefficient_Cp with a value around 0.03. We're not sure where this came from.

#We've now covered all equations up to eq 10 (p. 315)

##### EL CW Note #### 
#Equation 11 contains calculation of the metabolizable energy, but that doesn't seem to be used anywhere

#Now we calculate the ratio between net energy consumed and the digestible energy consumed, for maintenance (NE_DE) and growth (NEg_DE) (eqs. 12 and 13)

NE_DE<-(1.123-(4.092*(10^-3)*dig_energy)+
            (1.126*(10^-5)*(dig_energy^2))-
              (25.4/dig_energy))

NEg_DE<-(1.164-(5.160*(10^-3)*dig_energy)+
            (1.308*(10^-5)*(dig_energy^2))-
              (37.4/dig_energy))

##### EL CW Note #### 
#the following equation usually contains the energy mobilized by animals losing weight. This wasn't calculated, so we set this to 0
NE_mobilized<-0
#same for extra energy spent by working animals
NE_w<-0

gross_energy_intake<-(((NE_m+NE_mobilized+NE_activity+NE_lactation+NE_w+NE_pregnancy)/NE_DE)+
                          (NE_growth/NEg_DE))/(dig_energy/100)

enteric_CH4<-(gross_energy_intake*Y_m*365)/55.65

##### EL CW Note #### 
######Y_m  here is the methane conversion rate -
###### Joshua used the IPCC default value of 6.5%. Full story is on page 302 (a bit more complicated, with differentiation by animal type)

# Compute the CO2 equation for enteric methane
##### EL CW Note #### 
##### #This equation originally included multiplication by 365 - this seems to have been erroneous (already done in the previous line)
#This converts methane emissions to their Global Warming Potential (CO2-equivalent of CH4 is 25)
enteric_CO2<-enteric_CH4*25

# Manure management methane ($CH_4$) and nitrous oxide ($N_2O$)
# daily volatile solid excreted (page 10.42 of 2006 IPCC Guidelines for National Greenhouse Gas Inventories)

volsol<-(gross_energy_intake*
           (1-(dig_energy/100))+
           (urinary_energy*gross_energy_intake))*
             ((1-ash_content)/18.45)

@@

# the following is page 10.41 of the 2006 IPCC Guidelines for National Greenhouse Gas Inventories
mm_ch4<-(volsol*365)*
        (Bo*0.67*
           ((mms_pasture*MCFpasture)+
            (mms_spread*MCFspread)+
            (mms_drylot*MCFdrylot)+
            (mms_solidstore*MCFsolidstore)+
            (mms_composted*MCFcomposted)+
            (mms_liquid_slurry*MCFliquid)+
            (mms_biogas*MCFbiogas)+
            (mms_burn*MCFburn)+
              # the original code was lacking the 'sold' fraction - we added this for completeness
            (mms_sold*MCFsold)))

# Estimate CO2eq for manure methane
### ##### EL CW Note #### 
### this one also contained a multiplication by 365 that seemed erroneous
mm_CO2<-mm_ch4*25

@@

### ####
n_intake<-(gross_energy_intake/18.45)*((perc_crude_protein_diet/100)/6.25) # equation 10.32

# There was some code here on N retention in milk. That's apparently from equation 10.33, but it only had the first part of the equation, missing the part on weight gain. We decided to go with the alternative way of estimating N retention (figures given in Table 10.20)
# n_retention<-((milk_yield*(3.3/100))/6.38)

# Table 10.20 contains N retention coefficients:
# Dairy cows: 0.20 N_retention_dairy_cow
# Other cattle: 0.07 N_retention_other_cattle

animal_type_dairy_cow=1 ###ASSUMPTION

if(animal_type==animal_type_dairy_cow)
  n_excretion<-n_intake*(1-N_retention_dairy_cow)*365 else
    n_excretion<-n_intake*(1-N_retention_other_cattle)*365 #equation 10.31

# Direct nitrous oxide emissions (Eq 10.25 from IPCC guidelines)
mm_direct_n2o<-n_excretion*((mms_pasture*mndec_pasture)+
                           (mms_spread*mndec_spread)+
                           (mms_drylot*mndec_drylot)+
                           (mms_solidstore*mndec_solidstore)+
                           (mms_composted*mndec_composted)+
                           (mms_liquid*mndec_liquid)+
                           (mms_biogas*mndec_biogas)+
                           (mms_burn*mndec_burn)+
                           (mms_sold*mndec_sold))*(44/28)

# Nitrous oxide (N2O) from volatization (Eq 10.26 from IPCC guidelines)
mm_vol_n2o<-n_excretion*((mms_pasture*mnvc_pasture)+
                           (mms_spread*mnvc_spread)+
                           (mms_drylot*mnvc_drylot)+
                           (mms_solidstore*mnvc_solidstore)+
                           (mms_composted*mnvc_composted)+
                           (mms_liquid*mnvc_liquid)+
                           (mms_biogas*mnvc_biogas)+
                           (mms_burn*mnvc_burn)+
                           (mms_sold*mnvc_sold))*(44/28)

# Nitrous oxide (N2O) from leaching
# comment: this is probably wrong, since leached nitrogen doesn't emit (much) nitrous oxide
#mm_leach_n2o<-n_excretion*((mms_pasture*0.3)+(mms_spread*0.3)+(mms_drylot*0.3)+(mms_solidstore*0.3)+
#              (mms_composted*0.3)+(mms_liquid*0.3)+(mms_biogas*0)+(mms_burn*0))*0.0075*(44/28)

# Estimate CO2eq for manure nitrous
mm_CO2_N2Od<-mm_direct_n2o*298*365 # (298 is the CO2-equivalent of N2O)
mm_CO2_N2Ovol<-mm_vol_n2o*298*365
# mm_CO2_N2Oleach<-mm_leach_n2o*298*365
mm_CO2_N2O<-mm_CO2_N2Od+mm_CO2_N2Ovol #+mm_CO2_N2Oleach

@@

# Total emissions
on_farm<-enteric_CH4+enteric_CO2+mm_CO2+mm_CO2_N2O+feed_CO2

@@

return(list(enteric_CH4=enteric_CH4,
            #CW NOTE ####
            # ####working backwards from 'enteric_CO2' I see that 
            # #### this is the product of something twice multiplied by 365
            # ####### 
            milk_yield=milk_yield,
            enteric_CO2=enteric_CO2,
            mm_CO2=mm_CO2,
            mm_CO2_N2O=mm_CO2_N2O,
            feed_CO2=feed_CO2,
            on_farm=on_farm))
}

@@

```{r uncertainty_package}
uncertainty::varscatter(in_var = GHG_simulation$x$milk_yield, out_var = GHG_simulation$y$enteric_CH4)
```

>>>>

The formulas come from Chapter 10 of the *2006 IPCC Guidelines for National Greenhouse Gas Inventories* [@dong_ipcc_2006] (Volume 4 - Agriculture, Forestry and Other Land Use). The document can be found on [the IPCC website](https://www.ipcc-nggip.iges.or.jp/public/2006gl/pdf/4_Volume4/V4_10_Ch10_Livestock.pdf). The chapter is part of the book *Good Practice Guidance and Uncertainty Management in National Greenhouse Gas Inventories* [@ipcc_good_2021] , also on the [the IPCC website](https://www.ipcc-nggip.iges.or.jp/public/gp/english/). Many variables are based on the work done in *Variation in the carbon footprint of milk production on smallholder dairy farms in central Kenya* [@wilkes_variation_2020] and in *Methods and guidance to support MRV of livestock emissions* [@wilkes_methods_2019].

@@

# We start with the net energy for maintenance (NE_m; in MJ/day)
# This includes a maintenance coefficient, for which different values are recommended depending on animal type

# We're not sure what the animal types stand for, but this is important for calculating emissions.
# We need to know, which types are dairy cattle, which ones are lactating cows, which ones are bulls, and which ones are 'others'. We have 12 animal types.

animal_types_lact_cow<-c(4,5,6) # lactating cows - important to note here that the original data table appears to have worked with the actual number of months that cows were pregnant. This resulted in a weighted mean of the various maintenance coefficients. I'm not sure where we could get such data.

# animal type 1 - 'intact' grown-up males >=36 months
# animal type 2 - castrated grown-up males >=36 months
# animal type 3 - males 12-36 months
# animal type 4 - cows
# animal type 5 - cows
# animal type 6 - cows
# animal type 7 - females <= 12 months
# animal type 8 - males <= 12 months
# animal type 9 - females 12-36 months
# animal type 10 - female calves (<12 months)
# animal type 11 - male calves (<12 months)
# animal type 12 - first-time pregnant cows

# define C and Cfi based on sex (Page 10.17, below Equation 10.6)
if(animal_type %in% c(1,3,8,11)) growth_coefficient_C<-C_intact_males # males (1.2)
if(animal_type %in% c(2)) growth_coefficient_C<-C_castrates # castrates (1.0)
if(animal_type %in% c(4,5,6,7,9,10,12)) growth_coefficient_C<-C_females # females (0.8)

if(animal_type %in% c(4,5,6)) Cfi<-maintenance_coefficient_Cfi_lact_cow # lactating cows (0.386)
if(animal_type %in% c(1)) Cfi<-maintenance_coefficient_Cfi_bulls # bulls (0.37)
if(animal_type %in% c(2,3,7,8,9,10,11,12)) Cfi<-maintenance_coefficient_Cfi_other # others (0.322)

# define mature body weight (assuming this needs to be disaggregated by sex, for Eqation 10.6)
if(animal_type %in% c(1,2,3,8,11)) mature_weight_MW<-mature_weight_MW_male
if(animal_type %in% c(4,5,6,7,9,10,12)) mature_weight_MW<-mature_weight_MW_female

# define milk yield
if(animal_type %in% c(1,2,3,7,8,9,10,11,12)) milk_yield<-0
if(animal_type %in% c(4,5,6)) milk_yield<-eval(parse(text=paste0("milk_yield_",animal_type)))

#define pregnancy coefficient (full Cp for first-time pregnant, lower for later in most cases)
if(animal_type==12) Cp<-Cp_0
if(animal_type==4) Cp<-Cp_0*Cp_rate_4
if(animal_type==5) Cp<-Cp_0*Cp_rate_5
if(animal_type==6) Cp<-Cp_0*Cp_rate_6
if(animal_type %in% c(1,2,3,7,8,9,10,11)) Cp<-0

# define activity coefficient Ca (Table 10.5)
# Stall 0.00 - Animals are confined to a small area (i.e., tethered, pen, barn) with the result that they expend very little or no energy to acquire feed.
# Pasture 0.17 - Animals are confined in areas with sufficient forage requiring modest energy expense to acquire feed.
# Grazing large areas 0.36 - Animals graze in open range land or hilly terrain and expend significant energy to acquire feed.

# I don't quite understand how Ca was assigned in the original data table. There are many 0s for 'System 1' (Stall?), but also lots of values between 0 and 0.17 or 0.36. I can only imagine that these are weighted averages or reflect additional information about the feeding system, but this information doesn't seem to be given in the table.

if(feeding_system==1) C_activity<-Ca_stall # Stall (0.0)
if(feeding_system==2) C_activity<-Ca_pasture # Stall (0.17)
if(feeding_system==3) C_activity<-Ca_large_grazing # Stall (0.36)

# Mean live weight (kg) of animals surely varies by animal type, so we need separate estimates for them
live_weight_LW<-eval(parse(text=paste0("live_weight_LW_",animal_type)))

# Mean weight gain of animal type (kg/day)
weight_gain_WG<-eval(parse(text=paste0("weight_gain_WG_",animal_type)))

# Calculate the Net Energy for Maintenance (NE_m) (MJ/day) (Equation 10.3)

NE_m<-Cfi*(live_weight_LW^0.75)

# Calculate the Net Energy for Activity (NE_activity) (MJ/day) (Equation 10.4)

@@

#Net energy for growth (NE_growth) (MJ/day) (Equation 10.6)

NE_growth<-22.02*
  ((live_weight_LW/(growth_coefficient_C*mature_weight_MW))^0.75)*
  (weight_gain_WG^1.097)

@@

#Draft animal services are not accounted for
#NE_work <- 0.1 * NE_m * hours_per_day

#Energy need during pregnancy (Equation 10.13)
NE_pregnancy<-Cp*NE_m

# the Cp coefficient given in Equation 10.13 is 0.1, but in the sample this is only applied for animal type 12 (first-time pregnant). For all other lactating cows, this is reduced - not clear how this was done.

# Ratio of net energy available in a diet for maintenance to digestible energy consumed (Equation 10.14)
REM<-1.123-(4.092/1000*DE_perc)+(1.126/100000*DE_perc^2)-(25.4/DE_perc)

# Ratio of net energy available for growth in a diet to digestable energy consumed (Equation 10.15)
REG<-1.164-(5.160/1000*DE_perc)+(1.308/100000*DE_perc^2)-(37.4/DE_perc)

# Gross Energy (GE) MJ/day
NE_work<-0 # work animals aren't covered here, so we set this to zero
GE<-(((NE_m+NE_activity+NE_lactation+NE_work+NE_pregnancy)/REM)+(NE_growth/REG))/(DE_perc/100)

# CH4 emission factors for enteric fermentation from a livestock category (Equation 10.21)

enteric_CH4<-(GE*Y_m*365)/55.65 # kg CH4 per animal per year

# Compute the CO2-equivalent of methane emissions for enteric methane
#This converts methane emissions to their Global Warming Potential (CO2-equivalent of CH4 is 25)
enteric_CH4_CO2eq<-enteric_CH4*25

# Manure management methane ($CH_4$) and nitrous oxide ($N_2O$)
# daily volatile solid excreted (Equation 10.24)

volsol<-(GE*
           (1-(DE_perc/100))+
           (urinary_energy*GE))*
             ((1-ash_content)/18.45)

@@

# CH4 emission factor from manure management (Equation 10.23)
mm_ch4<-(volsol*365)*
        (Bo*0.67*
           ((mms_pasture*MCFpasture)+
            (mms_spread*MCFspread)+
            (mms_drylot*MCFdrylot)+
            (mms_solidstore*MCFsolidstore)+
            (mms_composted*MCFcomposted)+
            (mms_liquid_slurry*MCFliquid)+
            (mms_biogas*MCFbiogas)+
            (mms_burn*MCFburn)+
              # the original code was lacking the 'sold' fraction - we added this for completeness
            (mms_sold*MCFsold)))

# Estimate CO2eq for manure methane
mm_CH4_CO2eq<-mm_ch4*25

@@

# estimate N intake (kg/animal/day) (Equation 10.32)
N_intake<-(GE/18.45)*((perc_crude_protein_diet/100)/6.25)

# estimate N retention (kg/animal/day) (Equation 10.33)
N_retention<-((milk_yield*(1.9+0.4*milk_fat)/100)/6.38)+
  ((weight_gain_WG*(268-(7.03*NE_growth/weight_gain_WG)))/(1000*6.25))

# estimate N excretion (kg N/animal/yr) (Equation 10.31)
N_excretion<-N_intake*(1-N_retention)*365

# Direct nitrous oxide emissions (kg N2O/animal/yr) (Eq 10.25 from IPCC guidelines)
mm_direct_n2o<-N_excretion*((mms_pasture*mndec_pasture)+
                           (mms_spread*mndec_spread)+
                           (mms_drylot*mndec_drylot)+
                           (mms_solidstore*mndec_solidstore)+
                           (mms_composted*mndec_composted)+
                           (mms_liquid_slurry*mndec_liquid)+
                           (mms_biogas*mndec_biogas)+
                           (mms_burn*mndec_burn)+
                           (mms_sold*mndec_sold))*(44/28)

# N losses from volatilization (kg N/animal/yr) (Eq 10.26 from IPCC guidelines)
mm_vol_n<-N_excretion*((mms_pasture*mnvc_pasture)+
                           (mms_spread*mnvc_spread)+
                           (mms_drylot*mnvc_drylot)+
                           (mms_solidstore*mnvc_solidstore)+
                           (mms_composted*mnvc_composted)+
                           (mms_liquid_slurry*mnvc_liquid)+
                           (mms_biogas*mnvc_biogas)+
                           (mms_burn*mnvc_burn)+
                           (mms_sold*mnvc_sold))

# indirect N2O emissions due to volatilization of N from manure management (kg N2O/animal/year) (Equation 10.27)
mm_vol_n2o<-mm_vol_n*EF4*(44/28)

# Nitrous oxide (N2O) from leaching (kg N/animal/year) (equation 10.28)
mm_leach_n<-N_excretion*((mms_pasture*frac_leach_pasture)+
                           (mms_spread*frac_leach_spread)+
                           (mms_drylot*frac_leach_drylot)+
                           (mms_solidstore*frac_leach_solidstore)+
                           (mms_composted*frac_leach_composted)+
                           (mms_liquid_slurry*frac_leach_liquid)+
                           (mms_biogas*frac_leach_biogas)+
                           (mms_burn*frac_leach_burn)+
                           (mms_sold*frac_leach_sold))

# indirect N2O emissions due to leaching of N from manure management (kg N2O/animal/year) (Equation 10.29)
mm_leach_n2o<-mm_leach_n*EF5*(44/28)

# comment: this is probably wrong, since leached nitrogen doesn't emit (much) nitrous oxide
#mm_leach_n2o<-n_excretion*((mms_pasture*0.3)+(mms_spread*0.3)+(mms_drylot*0.3)+(mms_solidstore*0.3)+
#              (mms_composted*0.3)+(mms_liquid*0.3)+(mms_biogas*0)+(mms_burn*0))*0.0075*(44/28)

# Estimate CO2eq for manure nitrous
mm_N2O_CO2eq<-(mm_direct_n2o+mm_vol_n2o+mm_leach_n2o)*298 # (298 is the CO2-equivalent of N2O)

@@

# Total emissions
on_farm<-enteric_CH4_CO2eq+mm_CH4_CO2eq+mm_N2O_CO2eq+feed_CO2

@@

return(list(enteric_CH4=enteric_CH4,
            #CW NOTE ####
            # ####working backwards from 'enteric_CO2' I see that 
            # #### this is the product of something twice multiplied by 365
            # ####### 
            milk_yield=milk_yield,
            enteric_CH4_CO2eq=enteric_CH4_CO2eq,
            mm_CH4_CO2eq=mm_CH4_CO2eq,
            mm_N2O_CO2eq=mm_N2O_CO2eq,
            feed_CO2=feed_CO2,
            on_farm=on_farm))
}

@@

```{r uncertainty_package}
uncertainty::varscatter(in_var = GHG_simulation$x$milk_yield, out_var = GHG_simulation$y$enteric_CH4_CO2eq)
```
